---
title: "A3: Incarceration"
output: html_document
---

### Introduction

In the intricate tapestry of America's sociopolitical landscape, one of the most glaring manifestations of 
systemic racism is evident in its incarceration system. From overt policies to implicit biases, 
there has been a long-standing pattern of racial disparity in the country's jails and prisons.
Taking a deep dive into Washington's incarceration data, this report seeks to shed light on these disparities 
by analyzing the incarceration rates of Black and White populations in jails.
Two variables have been chosen for this analysis: black_jail_pop_rate representing the incarceration rate 
for Black individuals in jails, and white_jail_pop_rate for White individuals. 
These variables have been chosen due to the extensive discussions and documentation surrounding the racial 
disparities between Black and White populations in the U.S. incarceration system. By analyzing these metrics, 
I hope to draw attention to patterns that might otherwise go unnoticed, fostering informed discussions on this pressing issue.

### Summary Information

In the most recent year of our dataset (2018), Washington's incarceration data reveals a concerning disparity between Black and White populations.
On average, Black individuals face an incarceration rate in jails of 1052.113 per 100,000 people, 
as compared to White individuals who have a rate of 326.4833 per 100,000 people. 
When digging deeper into the counties, Lincoln County has the highest Black jail population rate, 
standing at 4918.03 per 100,000 people, while Whitman County has the lowest with 87.03 per 100,000 individuals.
Significantly, from 1990 to the latest year, there has been a change of -707.576 in the 
average jail population rate for Black individuals.

### The Dataset

#### Who collected the data?
  The data was collected by the Vera Institute. More specifically, it was drawn from
  the National Corrections Reporting Program (NCRP). In addition, data from the Annual Survey of Jails (ASJ),
  Census of Jails (COJ), and the Death in Custody Reporting Program (DCRP) were also used.

#### How was the data collected or generated?
  The data was collected by the prison/jail facilities that then report to these programs. The Vera institute then
  pulls it from there. 

##### Why was the data collected? 
  The data was collected in order to examine incarceration trends.

#### How many observations (rows) are in your data?
  1131 rows.
 
#### How many features (columns) are in the data?
  23 columns.

#### What, if any, ethical questions or questions of power do you need to consider when working with this data?
  The ethical issues I can think of when working with this data are:
  - Representation: Presenting jail rates, especially by racial or ethnic groups, can reinforce stereotypes. Interpret with care.
  - Data Collection: Was the collection consensual and ethical, respecting inmates' privacy rights?
  - Data Interpretation: Behind each number is a unique human story. Avoid dehumanizing data presentation.
  - Oversimplification: The data might not capture the whole story. Recognize its simplicity.
  - Use and Misuse: Ensure data presentation doesn't perpetuate discriminatory practices or policies.

#### What are possible limitations or problems with this data?   (at least 200 words)
There are many possible limitations or problems with this data. One that comes to mind is
that the data may exclude some jails or inmates, leading to incomplete representation. Human or systemic errors during data collection 
can affect accuracy and reliability. Additionally, since the data only reaches a specific year, it might not reflect current situations. 
Another issue is that we lack context: reasons for incarceration, sentence lengths, and confinement conditions remain unknown. 
The data's granularity might be insufficient, omitting factors like age or socio-economic status, which can provide deeper understanding.
External influences, such as legal changes or policing practices, can alter incarceration rates. Without considering these, interpretations can be biased. 
As the data pertains only to Washington, applying findings universally can be misleading. Inconsistent reporting across jails, like varying definitions of inmate counts,
can lead to discrepancies. In essence, while the dataset offers insights into incarceration,
it's imperative to understand its constraints and the broader implications that come with interpretation.

## Visualization

### Jail Population Rate Over Time

```{r chart1, echo=FALSE, warning=FALSE, message=FALSE}
# Libraries needed
library(tidyverse)
library(ggplot2)
library(viridis)

# Data import
url <- "https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-jail-rates-1990-WA.csv"
prison_dataset <- read.csv(url)

# Generate the line chart for jail population rate over time
ggplot(prison_dataset, aes(x = year)) +
  geom_line(aes(y = black_jail_pop_rate, color = "Black Jail Pop Rate"), na.rm = TRUE) +
  geom_line(aes(y = white_jail_pop_rate, color = "White Jail Pop Rate"), na.rm = TRUE) +
  labs(
    title = "Jail Population Rate Over Time for Black and White Inmates in Washington",
    x = "Year",
    y = "Jail Population Rate per 100,000",
    color = "Rate Type"
  ) +
  theme_minimal()
```

### Male vs Female Jail Population Rate

```{r chart2, echo=FALSE, warning=FALSE, message=FALSE}
# Generate scatter plot for male vs female jail population rate
ggplot(prison_dataset, aes(x = male_jail_pop_rate, y = female_jail_pop_rate)) +
  geom_point(aes(color = year), alpha = 0.6) +
  labs(
    title = "Comparison between Male and Female Jail Population Rate",
    x = "Male Jail Population Rate per 100,000",
    y = "Female Jail Population Rate per 100,000",
    color = "Year"
  ) +
  theme_minimal()
```

### Jail Population by County in Washington for 2018

```{r map, echo=FALSE, warning=FALSE, message=FALSE}
# Preprocess data for the map
us_county_map <- map_data("county")
washington_2018 <- subset(prison_dataset, year == 2018 & state == "WA")
washington_2018$cleaned_county_name <- tolower(gsub(" County", "", washington_2018$county_name))
washington_map <- subset(us_county_map, region == "washington")
combined_wa_data <- left_join(washington_map, washington_2018, by = c("subregion" = "cleaned_county_name"))

# Generate map
ggplot(data = combined_wa_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = total_jail_pop_rate), color = "white") +
  coord_fixed(1.3) + 
  scale_fill_viridis_c(name = "Total Jail\nPop Rate", 
                       na.value = "grey50",
                       guide = guide_colorbar(direction = "horizontal")) +
  labs(title = "Total Jail Population Rate by County in Washington (2018)",
       subtitle = "Grey areas represent missing data or non-matching county names") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```
